---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# multivarious

<!-- badges: start -->
<!-- badges: end -->

This is all very preliminary ... 

This package is intended to provide some basic abstractions and default implementations of basic computational infrastructure for multivariate component-based modeling such as principal components analysis.

The main idea is to model multivariate decompositions as involving projections from an input data space to a lower dimensional component space. This idea is encapsulated by the `projector` class and the `project` function. Support for two-way mapping (row projection and column projection) is provided by the derived class `bi-projector`. Generic functions for common operations are included:

- `project` for mapping from input space into (usually) reduced-dimensional output space
- `partial_project` for mapping a subset of input space into output space
- `project_vars` for mapping new variables ("supplementary variables") to output space
- `reconstruct` for reconstructing input data from its low-dimensional representation
- `residuals` for extracting residuals of a fit with `n` components.


## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("bbuchsbaum/multivarious")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(multivarious)
## basic example code
```

