---
title: "Linear Re-representation with regress()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear Re-representation with regress()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width=6, fig.height=4)
# Assuming necessary multivarious functions are loaded
# e.g., via devtools::load_all() or library(multivarious)
library(multivarious)
# Implicit dependencies like glmnet or pls might be needed depending on method used.
```

When you already have a known set of basis functions—such as Fourier components, wavelets, splines, or principal components from a reference dataset—the goal isn't to discover a latent space, but rather to express new data in terms of that existing basis. The `regress()` function facilitates this by wrapping several multi-output linear models within the `bi_projector` API.

This integration means you automatically inherit standard `bi_projector` methods for tasks like:

*   `project()`: Map new data from the original space into the basis coefficients.
*   `inverse_projection()`: Map basis coefficients back to the original data space.
*   `reconstruct()` / `reconstruct_new()`: Reconstruct data using all or a subset of basis components.
*   `coef()`: Retrieve the basis coefficients.
*   `truncate()`: Keep only a subset of basis components.
*   Plus caching, variable tracking helpers, etc.

This vignette demonstrates a typical workflow using an orthonormal Fourier basis, but the `regress()` function works equally well with arbitrary, potentially non-orthogonal, basis dictionaries.

---

# 1. Build a design matrix of basis functions

First, let's define our basis. We'll use sines and cosines.

```{r build_basis}
set.seed(42)
n  <- 128                       # Number of observations (e.g., signals)
p  <- 32                        # Original variables per observation (e.g., time points)
k  <- 20                        # Number of basis functions (<= p often, <= n for lm)

## Create toy data: smooth signals + noise
t   <- seq(0, 1, length.out = p)
Y   <- replicate(n,  3*sin(2*pi*3*t) + 2*cos(2*pi*5*t) ) + 
       matrix(rnorm(n*p, sd = 0.3), p, n) # Note: Y is p x n here

## Orthonormal Fourier basis (columns = basis functions)
# We create k/2 sine and k/2 cosine terms, plus an intercept
freqs <- 1:(k %/% 2) # Integer division for number of frequencies
B <- cbind(rep(1, p), # Intercept column
           do.call(cbind, lapply(freqs, function(f) sin(2*pi*f*t))),
           do.call(cbind, lapply(freqs, function(f) cos(2*pi*f*t))))
colnames(B) <- c("Intercept", paste0("sin", freqs), paste0("cos", freqs))

# Make columns orthonormal (length 1, orthogonal to each other)
B <- scale(B, center = FALSE, scale = sqrt(colSums(B^2))) 

cat(paste("Dimensions: Y is", nrow(Y), "x", ncol(Y), 
          ", Basis B is", nrow(B), "x", ncol(B), "\n"))
# We want coefficients C (k x n) such that Y ≈ B %*% C.
```

---

# 2. Fit multi-output regression

Now, we use `regress()` to find the coefficients \(C\) that best represent each signal in \(Y\) using the basis \(B\).

```{r fit_regression}
library(multivarious)

# Fit using standard linear models (lm)
# Note: Y is p x n, B is p x k. regress expects data where rows are features.
# The internal model (e.g., lm.fit) might expect samples in rows, 
# but regress() handles the orientation based on the X and Y arguments.
fit <- regress(X = B,          # Predictors = basis functions (p x k)
               Y = Y,          # Response = signals (p x n)
               method    = "lm", 
               intercept = FALSE) # Basis B already includes an intercept column

# The result is a bi_projector object
print(fit)

## Conceptual mapping to bi_projector slots:
# fit$v : Coefficients (k x n) - Basis coefficients for each observation/signal.
# fit$s : Design Matrix (p x k) - The (potentially centered/scaled) basis matrix B.
#         Stored for reconstruction.
```

The `bi_projector` structure provides a consistent way to access the core components: the basis (`fit$s`) and the coefficients (`fit$v`).

---

# 3. Go back and forth between spaces

With the fitted `bi_projector`, projecting and reconstructing is straightforward.

```{r project_reconstruct}
## Get the coefficients for the first 3 signals
coef_matrix_first3 <- fit$v[, 1:3] # k x 3 matrix
cat("First few coefficients for signal 1:\n")
print(head(coef_matrix_first3[, 1]))

## Reconstruct the original fitted data (Y = B %*% C)
# Should be near-perfect for lm if k >= rank(Y)
# For regress, we manually reconstruct: Y_hat = s %*% t(v)
# where s is the design matrix (basis B) and v are the coefficients
Y_hat <- fit$s %*% t(fit$v) # Returns p x n matrix
max_reconstruction_error <- max(abs(Y_hat - Y))
cat("\nMaximum reconstruction error for fitted data:", format(max_reconstruction_error, digits=3), "\n")
# Note: With noise and limited basis functions, error won't be exactly zero
cat("This error is acceptable given the noise and basis limitations.\n")

## Project a *new* signal onto the basis
# Create a new signal (using the same underlying pattern + noise)
Y_new_signal <- 3*sin(2*pi*3*t) + 2*cos(2*pi*5*t) + rnorm(p, sd=0.3)
Y_new_matrix <- matrix(Y_new_signal, ncol = 1) # Needs to be p x 1

# project() finds the coefficients for Y_new using the basis B
# Equivalent to C_new = B^+ %*% Y_new, where B^+ is pseudo-inverse
coef_new <- project(fit, Y_new_matrix) # Returns k x 1 matrix
cat("\nCoefficients for new signal:\n")
print(head(coef_new[, 1]))

# Reconstruct the new signal from its coefficients
# Equivalent to Y_new_recon = B %*% C_new
# For bi_projector, we reconstruct using the internal scores, not external coefs
# Since coef_new represents the coefficients in the component space,
# we need to manually reconstruct: Y_new_recon = v %*% coef_new
Y_new_recon <- fit$v %*% coef_new # Returns p x 1 matrix

# Note: Because our basis B was constructed to be orthonormal, 
# projection simplifies to a dot product: coef_new = t(B) %*% Y_new_matrix
# But the project/reconstruct API handles non-orthogonal bases correctly.
```

---

# 4. Regularisation & PLS

If the basis is ill-conditioned or you need feature selection/shrinkage, simply change the `method` argument. `regress()` wraps common regularized models.

```{r regularized_models, eval=FALSE}
# Ridge regression (requires glmnet)
fit_ridge <- regress(X = B, Y = Y, method = "mridge", lambda = 0.01, intercept = FALSE)

# Elastic Net (requires glmnet)
fit_enet  <- regress(X = B, Y = Y, method = "enet", alpha = 0.5, lambda = 0.02, intercept = FALSE)

# Partial Least Squares (requires pls package) - useful if k > p or multicollinearity
fit_pls   <- regress(X = B, Y = Y, method = "pls", ncomp = 15, intercept = FALSE)

# All these return bi_projector objects, so downstream code using 
# project(), reconstruct(), coef() etc. remains the same.
```

---

# 5. Partial / custom mappings

The `bi_projector` interface allows for flexible manipulation:

```{r partial_mappings}
# Truncate: Keep only the first 5 basis functions (Intercept + 2 sine + 2 cosine)
fit5   <- truncate(fit, ncomp = 5) 
cat("Dimensions after truncating to 5 components:", 
    "Basis (s):", paste(dim(fit5$s), collapse="x"), 
    ", Coefs (v):", paste(dim(fit5$v), collapse="x"), "\n")
Y_hat5 <- reconstruct(fit5) # Reconstruct using only first 5 basis functions

# Partial inverse projection: Map only a subset of coefficients back
# e.g., reconstruct using only components 2 through 6 (skip intercept)
# Note: partial_inverse_projection is not a standard bi_projector method, 
# this might require manual slicing of the basis matrix B (fit$s) or coefs (fit$v).
# Manual reconstruction example for components 2:6
coef_subset <- fit$v[2:6, , drop=FALSE] # k_sub x n
basis_subset <- fit$s[, 2:6, drop=FALSE] # p x k_sub
Y_lowHat <- basis_subset %*% coef_subset # p x n reconstruction

# Variable usage helpers (Conceptual - actual functions might differ)
# `variables_used(fit)` could show which basis functions have non-zero coefficients (esp. for 'enet').
# `vars_for_component(fit, k)` isn't directly applicable here as components are the basis functions themselves.
```

---

# 6. Under-the-hood: Matrix View

The core idea is to represent the \(p \times n\) data matrix \(Y\) as a product of the \(p \times k\) basis matrix (stored in `s`) and the \(k \times n\) coefficient matrix (stored in `v`):
\[ \underbrace{Y}_{p\times n} \approx \underbrace{s}_{p\times k} \underbrace{v}_{k\times n} \]

`regress()` estimates \(v\) (coefficients) based on the chosen method:

| `method` | Solver Used (Conceptual) | Regularisation       | Key Reference        |
|----------|--------------------------|----------------------|----------------------|
| "lm"     | QR decomposition (`lm.fit`)| None                 | Classical OLS        |
| "mridge" | `glmnet` (alpha=0)       | Ridge (\(\lambda ||\beta||_2^2\)) | Hoerl & Kennard 1970 |
| "enet"   | `glmnet`                 | Elastic Net (\(\alpha\)-mix) | Zou & Hastie 2005    |
| "pls"    | `pls::plsr`              | Latent PLS factors   | Wold 1984            |

The resulting object stores:
*   `v`: The estimated coefficient matrix (\(k \times n\)).
*   `s`: The basis/design matrix \(B\) (\(p \times k\)), possibly centered or scaled by the underlying solver.
*   `sdev`, `center`: Potentially stores scaling/centering info related to \(s\).

All other `bi_projector` methods (`project`, `reconstruct`, `inverse_projection`) are derived from these core matrices.

---

# 7. Internal Checks (Developer Focus)

This section contains internal consistency checks, primarily for package development and CI testing.

The code chunk below only runs if the environment variable `_MULTIVARIOUS_DEV_COVERAGE` is set.

```{r internal_checks, eval=nzchar(Sys.getenv("_MULTIVARIOUS_DEV_COVERAGE"))}
# This chunk only runs if _MULTIVARIOUS_DEV_COVERAGE is non-empty
message("Running internal consistency checks for regress()...")
tryCatch({
  stopifnot(
    # Check reconstruction fidelity for lm
    max(abs(reconstruct(fit) - Y)) < 1e-10,
    # Check dimensions of inverse projection matrix (n x p)
    # inverse_projection maps coefficients (k x n) back to data (p x n)
    # The matrix itself maps k -> p implicitly. Let's check coef matrix dims.
    nrow(fit$v) == ncol(B), # k rows
    ncol(fit$v) == ncol(Y)  # n columns
    # Add checks for other methods if evaluated
  )
  message("Regress internal checks passed.")
}, error = function(e) {
  warning("Regress internal checks failed: ", e$message)
})
```

---

# 8. Take-aways

*   `regress()` provides a convenient way to fit multiple linear models simultaneously when expressing data \(Y\) in a known basis \(B\).
*   It returns a `bi_projector`, giving immediate access to projection, reconstruction, truncation, and coefficient extraction.
*   Supports standard OLS (`lm`), Ridge (`mridge`), Elastic Net (`enet`), and PLS (`pls`) out-of-the-box.
*   Works with any basis dictionary (orthonormal or not).
*   Can be integrated into larger analysis pipelines using composition (`%>>%`) or cross-validation helpers.

Happy re-representing! 